### 2021-04-09 papers/thesis/full/ch2-spherical-harmonics/spherical-harmonics.tex
edits to SH

\subsection{Three-term recurrence relation for $\VSHvecfull(x,y,z)$}

@@

		\VSHvecfull_{l+1}(x,y,z) = -\Dlt (B_l-G_l(x,y,z)) \SHvec_l(x,y,z) - \Dlt C_l  \,\SHvec_{l-1}(x,y,z), \quad l \in \N.

@@

Let $\uvec$ be a vector-valued function on $\Omega$ whose values are of the form
\bseqn{
	\uvec(x,y,z) = u_\varphi(x,y,z) \: \phivec + u_\theta(x,y,z) \: \thetavec, \quad (x,y,z) \in \Omega
}
for some scalar functions $u^\varphi, u^\theta$ where $\phivec, \thetavec$ are the standard non-radial unit basis vectors on the sphere defined by $\phivec := (\cos \theta \sin \varphi, \sin \theta \sin \varphi, \cos \varphi)^\top$, $\thetavec := (-\sin \theta, \cos \theta, 0)^\top$. Such a function $\uvec$ we refer to as a \textit{vector-valued function in the tangent space of $\Omega$}. We can then approximate this function by its expansion in the VSH basis, i.e. for some large enough $N \in \N$:
\bseqn{
	\uvec(x,y,z) \approx \VSHvecfull(x,y,z)^\top \uvec^c := \sum^N_{l=0} \sum^l_{m=-l} \big[u^{\gradY}_{l,m} \, \gradYlm(x,y,z)  + u^{\gradpY}_{l,m} \, \gradpYlm(x,y,z) \big].
}

\subsubsection{Obtaining coefficients}
\bstodoinline{As for the scalar case, do I need to say more here about the transforms?}
Just like for the scalar case, we would like to be able to obtain the coefficients for the expansion of a vector valued function in the VSH basis up to a given degree. To do this, we can use a vector spherical harmonic transform. The development of vector spherical harmonic transforms is well established \cite{swarztrauber1993vector}, and continues today (see \cite{gia2019favest}).

\subsubsection{Function evaluation}
The Clenshaw algorithm presented in \bsrefsection{subsubsection:clenshaw} can also be used to evaluate a vector-valued function in the tangent space for a point $(x,y,z) \in \Omega$:
\bseqn{
	\quad &\text{1) } \text{Set } \vec{\xi}_{N+2} = \bold{0}, \: \vec{\xi}_{N+2} = \bold{0}. \\
	\quad &\text{2) } \text{For } n = N:-1:1 \\
	\quad & \quad \quad \quad \text{set } \vec{\xi}_{n}^\top = (\mathbf{u}^c_n)^\top - \vec{\xi}_{n+1}^\top \Dnt (B_n - G_n) -  \vec{\xi}_{n+2}^\top D^\top_{n+1} C_{n+1} \\
	\quad &\text{3) } \text{Output: } \uvec(x,y,z) \approx \VSHvecfull_1(x,y,z)^\top \vec{\xi}_{1} \\
	&\quad \quad \quad \quad \quad \quad \quad \quad \quad = (\vec{\xi}_{1})_1 \gradY_1^{-1} + (\vec{\xi}_{1})_2 \gradpY_1^{-1} + (\vec{\xi}_{1})_3 \gradY_1^{0} \\
	&\quad \quad \quad \quad \quad \quad \quad \quad \quad \quad \quad + (\vec{\xi}_{1})_4 \gradpY_1^{0} + (\vec{\xi}_{1})_5 \gradY_1^{1} + (\vec{\xi}_{1})_6 \gradpY_1^{1}.
}

\subsection{Sparse partial differential operators}

@@

In order to ensure we fully capitalise on the natural sparsity of the relationships so that we yield banded-block-banded operators, we will \enquote{split} coefficient vectors of a vector-valued function on the tangent space of $\Omega$ into two. On that note, define $\VSHvec, \VSHvecperp$ by
\bseqnnumber{
	\VSHvec &:= 
		\begin{pmatrix} 
			\VSHvec_0 \\ \VSHvec_1 \\ \vdots 
		\end{pmatrix}, \quad \text{where } \quad 
	\VSHvecl := 
		\begin{pmatrix} 
			(\gradY^{-l}_l)^\top \\ \vdots \\ (\gradY^{l}_l)^\top
		\end{pmatrix} \in \C^{(2l+1) \times 3} \quad \forall \: l \in \No, \label{eqn:VSHdef} \\
	\VSHvecperp &:= 
		\begin{pmatrix} 
			\VSHvecperp_0 \\ \VSHvecperp_1 \\ \vdots 
		\end{pmatrix}, \quad \text{where } \quad 
	\VSHvecperpl := 
		\begin{pmatrix} 
			(\gradpY^{-l}_l)^\top \\ \vdots \\ (\gradpY^{l}_l)^\top
		\end{pmatrix} \in \C^{(2l+1) \times 3} \quad \forall \: l \in \No. \label{eqn:VSHperpdef}
}
Let $\uvec$ be a vector-valued function on the tangent space of $\Omega$ with coefficients vectors $\uvecc, \uveccperp$ for its expansion in the VSH basis up to order $N \in \N$, and let $f$ be a scalar function on $\Omega$ with coefficients vector $\fvecc$ for its expansion in the SSH basis also up to order $N$, i.e.
\bseqn{
	\uvec(x,y,z) &\approx \VSHvec(x,y,z)^\top \: \uvecc + \VSHvecperp(x,y,z)^\top \: \uveccperp \\
	&:=  \sum^N_{l=0} \sum^l_{m=-l} \big[u^{\gradY}_{l,m} \, \gradYlm(x,y,z)  + u^{\gradpY}_{l,m} \, \gradpYlm(x,y,z) \big], \\
	f(x,y,z) &\approx \SHvec(x,y,z)^\top \: \fvecc :=  \sum^N_{l=0} \sum^l_{m=-l} f_{l,m} \, \Ylm(x,y,z)
}
We are now in a position to derive the sparse differential operators that can be applied to the functions $\uvec, f$.

\begin{definition}\label{def:SHoperators}
	We define the operator matrices $\cal{D}, \calG, \calL$ according to:
\bseqn{
	\divergence \uvec(x,y,z) &= \SHvec(x,y,z)^\top \: \calD \: \uvecc, \\
	\gradS f(x,y,z) &= \VSHvec(x,y,z)^\top \: \calG \: \fvecc, \\
	\DeltaS f(x,y,z) &= \SHvec(x,y,z)^\top \: \calL \: \fvecc.
}
\end{definition}

\begin{theorem}
	The operator matrices defined in \bsrefdef{def:SHoperators} are diagonal, and are given by:
\bseqn{
	\calG &= I_{(N+1)^2}, \quad
	\calD \equiv \calL := 
		\begin{pmatrix}
			\calL_0 & & \\
			& \ddots & \\
			& & \calL_N
		\end{pmatrix},
}
where
\bseqn{
	\calL_l &:= 
		\begin{pmatrix}
			-l(l+1) & & \\
			& \ddots & \\
			& & -l(l+1)
		\end{pmatrix} \in \R^{(2l+1) \times (2l+1)}.
}
\end{theorem}

\begin{proof}
	The argument for $\calG$ is trivial by definition. Now, the spherical harmonics, as the name eludes to, satisfy a harmonic relationship $\DeltaS \Ylm = \: l(l+1) \Ylm$ (see e.g. \cite{humi1987factorisation}). Thus, for the proof for $\calD$, consider
\bseqn{
	\divergence \uvec &= \divergence \Big( \sum_{l=0}^N \sum_{m=-l}^{l} [ u^{\gradY}_{l,m} \gradYlm + u^{\gradpY}_{l,m} \gradpYlm ] \Big) \\
	&= \sum_{l=0}^N \sum_{m=-l}^{l} u^{\gradY}_{l,m} \: \DeltaS \Ylm \\
	&= \sum_{l=0}^N \sum_{m=-l}^{l} - u^{\gradY}_{l,m} \: l(l+1) \Ylm,
}
using the fact that $\divergence (\grad^\perp f )\equiv 0$ for any function $f$. We similarly have for $\calL$:
\bseqn{
	\DeltaS f &= \DeltaS \Big( \sum_{l=0}^N \sum_{m=-l}^{l} f_{l,m} \Ylm \Big) \\
	&= \sum_{l=0}^N \sum_{m=-l}^{l} f_{l,m} \: \DeltaS \Ylm \\
	&= \sum_{l=0}^N \sum_{m=-l}^{l} - f_{l,m} \: l(l+1) \Ylm.
}
\end{proof}

@@

Let $\uvec(x,y,z)$ be the tangential velocity of a flow and $h(x,y,z)$ be the height deviation of the flow from some constant reference height $\calH$. Define $\rvec$ as the unit outward normal vector at the point on the sphere $(x,y,z)$, so that
\bseqn{
	\rvec = \begin{pmatrix} x \\ y \\ z \end{pmatrix}.
}
The linear SWEs are
\bseqn{
	\begin{cases}
		\pfpx{t}{\uvec} + f \rvec \times \uvec - \nabla h = \mathbf{0} \\
		\pfpx{t}{h} + \calH \nabla \cdot \uvec = 0
	\end{cases}
}
where $f = 2 \rotrate \cos(\varphi) =  2 \rotrate z$ is the Coriolis parameter, and $\rotrate = 7.2921 × 10^{-5}$ rad s$^{-1}$ is the rotation rate of the Earth (note that $\varphi = {\pi \over 2} - \text{latitude}$).

A keen eye may notice that we require one more operator not yet defined -- one for the Coriolis parameter cross product (i.e. $f \: \rvec \times \uvec$). 
\begin{definition}\label{def:coriolisoperator}
	Define the operator matrices $\calF, \calF^\perp$ according to:
\bseqn{
	f \: \rvec \times \uvec(x,y,z) &= \VSHvec(x,y,z)^\top \: \calF \: \uveccperp + \Big(\VSHvecperp(x,y,z)\Big)^\top \: \calF^\perp \: \uvecc
}
\end{definition}
\begin{lemma}
	The operators $\calF, \calF^\perp$ defined in \bsrefdef{def:coriolisoperator} are sparse with banded-block-banded structure. More specifically,
\begin{itemize}
	\item $\calF$ has block-bandwidths $(1,1)$, and sub-block-bandwidths $(1,1)$
	\item $\calF^\perp$ has block-bandwidths $(1,1)$, and sub-block-bandwidths $(1,1)$
\end{itemize}
\end{lemma}
\begin{proof}
	Using \bsreflemma{lemma:VSHrecurrences}, we have that:
\bseqn{
	f \: \rvec \times \bold{u} &= f \: \rvec \times \sum_{l=0}^{N} \sum_{m=-l}^{l} [ u^{\gradY}_{l,m} \gradYlm + u^{\gradpY}_{l,m} \gradpYlm ] \\
	&= 2 \: \rotrate \: z \: \sum_{l=0}^{N} \sum_{m=-l}^{l} [ u^{\gradY}_{l,m} \rvec \times \gradYlm + u^{\gradpY}_{l,m} \rvec \times ( \rvec \times \gradYlm ) ] \\
	&= 2 \: \rotrate \: \sum_{l=0}^{N} \sum_{m=-l}^{l} [ u^{\gradY}_{l,m} z \: \gradpYlm - u^{\gradpY}_{l,m} z \: \gradYlm ) ] \\
	&= 2 \: \rotrate \: \sum_{l=0}^{N} \sum_{m=-l}^{l} \Big[ u^{\gradY}_{l,m} \: \Big( \Gamma_{l,m,1}^* \gradpY^{m}_{l-1} + \Gamma_{l,m,3}^* \gradpY^{m}_{l+1} + \Gamma_{l,m,2} \gradY^{m}_{l} \Big) \\
	&\quad \quad \quad \quad \quad \quad \quad \quad - u^{\gradpY}_{l,m} \: \Big( \Gamma_{l,m,1} \gradY^{m}_{l-1} + \Gamma_{l,m,3} \gradY^{m}_{l+1} + \Gamma_{l,m,2}^* \gradpY^{m}_{l} \Big) \Big].
}
We can then write down the operators $\calF, \calF^\perp$ as follows:
\bseqn{
	\calF &= \bar{\calF^\perp} = 2 \: \rotrate \: 
		\begin{pmatrix}
			\calF_{0,2} & \calF_{1,1} & & & & \\
			\calF_{0,3} & \calF_{1,2} & \calF_{2,1} & & & \\
			& \calF_{1,3} & \calF_{2,2} & \calF_{3,1}  & & & \\
			& & \calF_{2,3} & \ddots & \ddots & \\
			& & & \ddots & \ddots & \ddots \\
		\end{pmatrix}
}
where $\bar{A}$ denotes the conjugate matrix of $A$ (i.e. where the entries of $\bar{A}$ are simply the complex conjugates of the entries of $A$) \cite[p.10]{courant1953methods}, and for $l = 0,\dots,N$
\bseqn{
	\calF_{l,1} &:= 
		\begin{pmatrix}
			0 & \Gamma_{l,-l+1,1} & & \\
			& & \ddots & & \\
			& & & \Gamma_{l,l-1,1} & 0 \\
		\end{pmatrix} \in \R^{(2l-1)\times(2l+1)}, \quad (l \ne 0), \\
	\calF_{l,2} &:= 
		\begin{pmatrix}
			\Gamma_{l,-l,2} & & \\
			& \ddots & \\
			& & \Gamma_{l,l,2} \\
		\end{pmatrix}  \in \R^{(2l+1)\times(2l+1)}, \\
	\calF_{l,3} &:= 
		\begin{pmatrix}
			0 \\
			\Gamma_{l,-l,3} \\
			& & \ddots \\
			& & & & \Gamma_{l,l,3} \\
			& & & & 0 \\
		\end{pmatrix} \in \R^{(2l+3)\times(2l+1)}.
}
\end{proof}

For simplicity, we implement a backward Euler timestepping method to solve the linear SWEs with timestep $\Delta t$:
\bseqn{
	\uvec_{n+1} &= \uvec_{n} + \Delta t \: (\nabla h_{n+1} - f \rvec \times \uvec_{n+1}) \\
	h_{n+1} &= h_n - \Delta t \: \calH \nabla \cdot \uvec_{n+1}
}
In coefficient space, this system becomes:
\bseqn{
	\uvecc_{n+1} &= \uvecnc + \Delta t \: (\calG \vec{h}^c_{n+1} - \calF \uveccperp_{n+1}) \\
	\uveccperp_{n+1} &=  - \Delta t \: \calF^\perp \uvecc_{n+1} \\
	\vec{h}^c_{n+1} &= \vec{h}^c_n - \Delta t \: \calH D \uvecc_{n+1}
}
$\iff$
\bseqn{
	\uvecc_{n+1} &= \uvecc_{n} + \Delta t \: (\calG \vec{h}^c_n - \Delta t \: \calH \calG \calD \uvecc_{n+1} + \Delta t \calF \calF^\perp \uvecc_{n+1}) \\
	\uveccperp_{n+1} &=  - \Delta t \: \calF^\perp \uvecc_{n+1} \\
	\vec{h}^c_{n+1} &= \vec{h}^c_n - \Delta t \: \calH D \uvecc_{n+1}
}
$\iff$
\bseqn{
	\uvecc_{n+1} &= (I + \Delta t^2 \calH \calG \calD - \Delta t^2 \calF \calF^\perp)^{-1} \: (\uvecc_{n} + \Delta t \: \calG \vec{h}^c_n) \\
	\uveccperp_{n+1} &=  - \Delta t \: \calF^\perp \uvecc_{n+1} \\
	\vec{h}^c_{n+1} &= \vec{h}^c_n - \Delta t \: \calH \calD \uvecc_{n+1}.
}
Here, it is clear that the matrix $I + \Delta t^2 \calH \calG \calD - \Delta t^2 \calF \calF^\perp$ is sparse with banded-block-banded structure, and hence this system will be efficient to solve. Doing so at each iteration provides us with the coefficients vectors $\uvecc_{n+1}, \uveccperp_{n+1}, \vec{h}^{c}_{n+1}$ if we know $\uvecc_{n}, \uveccperp_{n}, \vec{h}^{c}_{n}$.

\bstodo{It would be nice to have some images here. However I need to update/fix/tidy my SH code as its probably got a few bugs in it having now gone back over the derivations that I have done.}

>>>>

\subsection{Three-term recurrence relation for $\VSHvecfull(x,y,z)$}\label{subsection:VSHrecurrence}

@@

		\VSHvecfull_{l+1}(x,y,z) = -\Dlt (B_l-G_l(x,y,z)) \VSHvecfull_l(x,y,z) - \Dlt C_l  \, \VSHvecfull_{l-1}(x,y,z), \quad l \in \N.

@@

It will be useful to be able to describe the coefficients of a vector-valued function on the tangent space of $\Omega$ in multiple ways, and as such we will introduce some more notation for the \enquote{splitting} of the basis vectors into the two groups stemming from the gradient and perpendicular gradient directions. Let's explain what we mean by this. Let $\uvec$ be a vector-valued function on $\Omega$ whose values are of the form
\bseqn{
	\uvec(x,y,z) = u_\varphi(x,y,z) \: \phivec + u_\theta(x,y,z) \: \thetavec, \quad (x,y,z) \in \Omega
}
for some scalar functions $u^\varphi, u^\theta$ where $\phivec, \thetavec$ are the standard non-radial unit basis vectors on the sphere defined by $\phivec := (\cos \theta \sin \varphi, \sin \theta \sin \varphi, \cos \varphi)^\top$, $\thetavec := (-\sin \theta, \cos \theta, 0)^\top$. Such a function $\uvec$ we refer to as a \textit{vector-valued function in the tangent space of $\Omega$}.

Define $\VSHvec, \VSHvecperp$ by
\bseqnnumber{
	\VSHvec &:= 
		\begin{pmatrix} 
			\VSHvec_0 \\ \VSHvec_1 \\ \vdots 
		\end{pmatrix}, \quad \text{where } \quad 
	\VSHvecl := 
		\begin{pmatrix} 
			(\gradY^{-l}_l)^\top \\ \vdots \\ (\gradY^{l}_l)^\top
		\end{pmatrix} \in \C^{(2l+1) \times 3} \quad \forall \: l \in \No, \label{eqn:VSHdef} \\
	\VSHvecperp &:= 
		\begin{pmatrix} 
			\VSHvecperp_0 \\ \VSHvecperp_1 \\ \vdots 
		\end{pmatrix}, \quad \text{where } \quad 
	\VSHvecperpl := 
		\begin{pmatrix} 
			(\gradpY^{-l}_l)^\top \\ \vdots \\ (\gradpY^{l}_l)^\top
		\end{pmatrix} \in \C^{(2l+1) \times 3} \quad \forall \: l \in \No. \label{eqn:VSHperpdef}
}
If $\uvec$ is a vector-valued function on the tangent space of $\Omega$, it will have the coefficients vectors $\uvec^c, \uvecc, \uveccperp$ for its expansion in the VSH basis up to order $N \in \N$, defined such that
\bseqn{
	\uvec(x,y,z) &\approx \VSHvecfull(x,y,z)^\top \: \uvec^c \\
	&\equiv \VSHvec(x,y,z)^\top \: \uvecc + \VSHvecperp(x,y,z)^\top \: \uveccperp \\
	&:=  \sum^N_{l=0} \sum^l_{m=-l} \big[u^{\gradY}_{l,m} \, \gradYlm(x,y,z)  + u^{\gradpY}_{l,m} \, \gradpYlm(x,y,z) \big].
}
Of course, just like for the scalar case, we can use transforms to compute the coefficients of such a known function, and further evaluate, and apply differential operators to, a function knowing just its coefficients.

\subsubsection{Obtaining coefficients}
\bstodoinline{As for the scalar case, do I need to say more here about the transforms?}
As eluded to, we would like to be able to obtain the coefficients for the expansion of a vector valued function in the VSH basis up to a given degree, similar to how we approach the scalar case. To do this, we can use a vector spherical harmonic transform. The development of vector spherical harmonic transforms is well established \cite{swarztrauber1993vector}, and continues today (see \cite{gia2019favest}).

\subsubsection{Function evaluation}
The Clenshaw algorithm presented in \bsrefsection{subsubsection:clenshaw} can also be used to evaluate a vector-valued function in the tangent space for a point $(x,y,z) \in \Omega$, where the matrices $C_n, B_n, \Dnt, G_n(x,y,z)$ are all defined in \bsrefsection{subsection:VSHrecurrence}:
\bseqn{
	\quad &\text{1) } \text{Set } \vec{\xi}_{N+2} = \bold{0}, \: \vec{\xi}_{N+2} = \bold{0}. \\
	\quad &\text{2) } \text{For } n = N:-1:1 \\
	\quad & \quad \quad \quad \text{set } \vec{\xi}_{n}^\top = (\uvec^c_n)^\top - \vec{\xi}_{n+1}^\top \Dnt \big(B_n - G_n(x,y,z)\big) -  \vec{\xi}_{n+2}^\top D^\top_{n+1} C_{n+1} \\
	\quad &\text{3) } \text{Output: } \uvec(x,y,z) \approx \VSHvecfull_1(x,y,z)^\top \vec{\xi}_{1} \\
	&\quad \quad \quad \quad \quad \quad \quad \quad \quad = (\vec{\xi}_{1})_1 \gradY_1^{-1} + (\vec{\xi}_{1})_2 \gradpY_1^{-1} + (\vec{\xi}_{1})_3 \gradY_1^{0} \\
	&\quad \quad \quad \quad \quad \quad \quad \quad \quad \quad \quad + (\vec{\xi}_{1})_4 \gradpY_1^{0} + (\vec{\xi}_{1})_5 \gradY_1^{1} + (\vec{\xi}_{1})_6 \gradpY_1^{1}.
}
Here, at each iteration, the vector $\uvec^c_n$ is given by the merging of the two order $n$ subvector of $\uvec^c$:
\bseqn{
	\uvec^c_n := 
		\begin{pmatrix} 
			u^{\gradY}_{n,-n} \\
			u^{\gradpY}_{n,-n} \\
			\vdots \\
			u^{\gradY}_{n,n} \\
			u^{\gradpY}_{n,n} \\
		\end{pmatrix}.
}
Notice how the iteration also ends at $n=1$ (not $n=0$) here, since the VSHs in our tangent space basis (and hence coefficients) for degree $n=0$ are zero.

\subsection{Sparse partial differential operators}\label{section:SHdiffoperators}

@@

Let $\uvec$ be a vector-valued function on the tangent space of $\Omega$ with coefficients vectors $\uvecc, \uveccperp$ for its expansion in the VSH basis up to order $N \in \N$, and let $f$ be a scalar function on $\Omega$ with coefficients vector $\fvecc$ for its expansion in the SSH basis also up to order $N$, i.e.
\bseqn{
	\uvec(x,y,z) &\approx \VSHvecfull(x,y,z)^\top \: \uvec^c \\
	&\equiv \VSHvec(x,y,z)^\top \: \uvecc + \VSHvecperp(x,y,z)^\top \: \uveccperp \\
	&:=  \sum^N_{l=0} \sum^l_{m=-l} \big[u^{\gradY}_{l,m} \, \gradYlm(x,y,z)  + u^{\gradpY}_{l,m} \, \gradpYlm(x,y,z) \big], \\
	f(x,y,z) &\approx \SHvec(x,y,z)^\top \: \fvecc :=  \sum^N_{l=0} \sum^l_{m=-l} f_{l,m} \, \Ylm(x,y,z).
}
In order to fully exploit the sparsity of the relationships and ensure we are using banded-block-banded operators, in problems where we have an interdependency of scalar and vector-valued functions it can be useful to \enquote{expand} the scalar function's coefficients vector $\fvecc$ to be the length of the vector-valued one $\uvec^c$, with zeros being inserted after each entry, i.e. define the \textit{expanded coefficients for a scalar function on $\Omega$ in the SH basis} by
\bseqn{
	\fveccexp :=
		\begin{pmatrix}
			\fveccexp_0 \\
			\vdots \\
			\fveccexp_N
		\end{pmatrix} \in \R^{2(N+1)^2}, \quad
	\fveccexp_l := 
		\begin{pmatrix}
			f_{l,-l} \\
			0 \\
			f_{l,-l+1} \\
			0 \\
			\vdots \\
		 	f_{l,l} \\
			0
		\end{pmatrix} \in \R^{4(l+1)}.
}
To help with the definitions of the operators we will discuss, let $\calE_s, \calE_v$ be (severely) rectangular matrices defined such that $\calE_s \fveccexp = \fvecc$ and $\calE_v \fvecc = \fveccexp$. We are now in a position to derive the sparse differential operators that can be applied to the functions $\uvec, f$.

\begin{definition}\label{def:SHoperators}
	We define the operator matrices $\calD, \tilde \calD, \calG, \tilde \calG \calL$ according to:
\bseqn{
	\divergence \uvec(x,y,z) &= \SHvec(x,y,z)^\top \: \calD \: \uvecc, \\
	\divergence \uvec(x,y,z) &= \SHvec(x,y,z)^\top \: \calE_s \: \big(\tilde \calD \: \uvec^c), \\
	\gradS f(x,y,z) &= \VSHvec(x,y,z)^\top \: \calG \: \fvecc, \\
	\gradS f(x,y,z) &= \VSHvecfull(x,y,z)^\top \: \tilde \calG \: \fveccexp, \\
	\DeltaS f(x,y,z) &= \SHvec(x,y,z)^\top \: \calL \: \fvecc.
}
\end{definition}
\begin{remark}
	The operators $\tilde \calD, \tilde \calG$ would be used when vectors of the resulting length are required for the problem you are wishing to solve.
\end{remark}

\begin{theorem}
	The operator matrices defined in \bsrefdef{def:SHoperators} are diagonal, and are given by:
\bseqn{
	\calD &\equiv \calL := 
		\begin{pmatrix}
			\calL_0 & & \\
			& \ddots & \\
			& & \calL_N
		\end{pmatrix} \in \R^{(N+1)^2}, \quad
	\tilde \calD = 
		\begin{pmatrix}
			\calL_0 & & & & \\
			& 0 & & & \\
			& & \ddots & & \\
			& & & \calL_N & \\
			& & & & 0
		\end{pmatrix} \in \R^{2(N+1)^2}, \\
	\calG &= I_{(N+1)^2}, \quad
	\tilde \calG = 
		\begin{pmatrix}
			1 & & & & \\
			& 0 & & & \\
			& & \ddots & & \\
			& & & 1 & \\
			& & & & 0
		\end{pmatrix}  \in \R^{2(N+1)^2},
}
where $\calL_l := -l(l+1) \: I_{2l+1}$.
\end{theorem}

\begin{proof}
	The arguments for $\calG$ and $\tilde \calG$ are trivial by definition. The spherical harmonics, as the name eludes to, satisfy a harmonic relationship $\DeltaS \Ylm = \: l(l+1) \Ylm$ (see e.g. \cite{humi1987factorisation}). Thus, for the proof for $\calD$ and $\tilde \calD$, consider
\bseqn{
	\divergence \uvec &= \divergence \Big( \sum_{l=0}^N \sum_{m=-l}^{l} [ u^{\gradY}_{l,m} \gradYlm + u^{\gradpY}_{l,m} \gradpYlm ] \Big) \\
	&= \sum_{l=0}^N \sum_{m=-l}^{l} u^{\gradY}_{l,m} \: \DeltaS \Ylm \\
	&= \sum_{l=0}^N \sum_{m=-l}^{l} - u^{\gradY}_{l,m} \: l(l+1) \Ylm,
}
using the fact that $\divergence (\grad^\perp f )\equiv 0$ for any function $f$. We similarly have for $\calL$:
\bseqn{
	\DeltaS f &= \DeltaS \Big( \sum_{l=0}^N \sum_{m=-l}^{l} f_{l,m} \Ylm \Big) \\
	&= \sum_{l=0}^N \sum_{m=-l}^{l} f_{l,m} \: \DeltaS \Ylm \\
	&= \sum_{l=0}^N \sum_{m=-l}^{l} - f_{l,m} \: l(l+1) \Ylm.
}
\end{proof}

@@

Let $\uvec(x,y,z)$ be the tangential velocity of a flow and $h(x,y,z)$ be the height deviation of the flow from some constant reference height $\calH$. Define $\rvec$ as the unit outward normal vector at the point on the sphere $(x,y,z)$, so that
\bseqn{
	\rvec = \begin{pmatrix} x \\ y \\ z \end{pmatrix}.
}
The linear SWEs are
\bseqn{
	\begin{cases}
		\pfpx{t}{\uvec} + f \rvec \times \uvec - \gradS h = \mathbf{0} \\
		\pfpx{t}{h} + \calH \gradS \cdot \uvec = 0
	\end{cases}
}
where $f = 2 \rotrate \cos(\varphi) =  2 \rotrate z$ is the Coriolis parameter, and $\rotrate$ is the rotation rate of the Earth (note that $\varphi = {\pi \over 2} - \text{latitude}$ \bstodo{make a footnote}). On a full scale Earth, the rotation rate is $\rotrate = 7.2921 × 10^{-5}$ rad s$^{-1}$ \bstodo{make a footnote}.

A keen eye may notice that -- while we have operators for the multiplication by z, the spherical gradient and divergence -- we require one more operator not yet defined for the unit vector cross product $\rvec \times \uvec$.
\begin{definition}\label{def:unitveccrossproduct}
	Define the operator matrix $\calR$ according to:
\bseqn{
	\rvec(x,y,z) \times \uvec(x,y,z) &= \VSHvecfull(x,y,z)^\top \: \calR \uvec^c.
}
\end{definition}
\begin{lemma}
	The operator $\calR$ defined in \bsrefdef{def:unitveccrossproduct} is sparse with banded-block-banded structure. More specifically, $\calR$ has block-bandwidths $(0,0)$, and sub-block-bandwidths $(1,1)$ (i.e. is tridiagonal).
\end{lemma}
\begin{proof}
	Using the relations that $\gradYlm = \gradS \Ylm$ and $\gradpYlm = \rvec \times \gradS \Ylm$ for each $l \in \N$, $m \in \Z$ s.t. $-l \le m \le l$, we have that:
\bseqn{
	\rvec \times \bold{u} &= \rvec \times \sum_{l=0}^{N} \sum_{m=-l}^{l} [ u^{\gradY}_{l,m} \gradYlm + u^{\gradpY}_{l,m} \gradpYlm ] \\
	&= \sum_{l=0}^{N} \sum_{m=-l}^{l} [ u^{\gradY}_{l,m} \rvec \times \gradYlm + u^{\gradpY}_{l,m} \rvec \times ( \rvec \times \gradYlm ) ] \\
	&= \sum_{l=0}^{N} \sum_{m=-l}^{l} [ u^{\gradY}_{l,m} \: \gradpYlm - u^{\gradpY}_{l,m} \: \gradYlm ) ].
}
We can then write down the operator $\calR$ as follows:
\bseqn{
	\calR &=
		\begin{pmatrix}
			R & & & \\
			& R & & \\
			& & & \ddots
		\end{pmatrix} \in \R^{2(N+1)^2 \times 2(N+1)^2},
}
where the blocks $R$ are simply given by
\bseqn{
	R &:= 
		\begin{pmatrix}
			0 & 1 \\
			-1 & 0
		\end{pmatrix} \in \R^{2\times2}.
}
\end{proof}

For simplicity, we will implement a backward Euler timestepping method to solve the linear SWEs with timestep $\Delta t$:
\bseqn{
	\uvec_{n+1} &= \uvec_{n} + \Delta t \: (\gradS h_{n+1} - f \rvec \times \uvec_{n+1}) \\
	h_{n+1} &= h_n - \Delta t \: \calH \gradS \cdot \uvec_{n+1}
}
Let's write this in coefficient space. Firstly, our coefficients vectors for the functions $\uvec, h$ we will denote $\uvec^c, \vec{h}^c$ -- that is:
\bseqn{
	\uvec(x,y,z) &\approx \VSHvecfull(x,y,z)^\top \: \uvec^c \\
	h(x,y,z) &\approx \SHvec(x,y,z)^\top \: \vec{h}^c.
}
As eluded to earlier, we will \enquote{extend} the scalar function's coefficients vector, writing $\tilde{\vec{h}}^c = E_v \vec{h}^c$, as described in \bsrefsection{section:SHdiffoperators}. Thus, our system can be written as a matrix-vector system:
\bseqn{
	\uvec^c_{n+1} &= \uvec^c_n + \Delta t \: (\tilde \calG \tilde{\vec{h}}^c_{n+1} - 2 \rotrate J_z \calR \uvec^c_{n+1}) \\
	\tilde{\vec{h}}^c_{n+1} &= \tilde{\vec{h}}^c_n - \Delta t \: \calH \tilde \calD \uvec^c_{n+1}
}
$\iff$
\bseqn{
	\uvec^c_{n+1} &= \uvec^c_n + \Delta t \: (\tilde \calG \tilde{\vec{h}}^c_{n} - \Delta t \: \calH \tilde \calG \tilde \calD \uvecc_{n+1} - 2 \rotrate J_z \calR \uvec^c_{n+1}) \\
	\tilde{\vec{h}}^c_{n+1} &= \tilde{\vec{h}}^c_n - \Delta t \: \calH \tilde \calD \uvec^c_{n+1}
}
$\iff$
\bseqn{
	\Big[ I + \Delta t^2 \calH \tilde \calG \tilde \calD - 2 \rotrate \Delta t \: J_z \calR \Big] \: \uvec^c_{n+1} &= \uvec^c_n + \Delta t \tilde \calG \tilde{\vec{h}}^c_{n} \\
	\tilde{\vec{h}}^c_{n+1} &= \tilde{\vec{h}}^c_n - \Delta t \: \calH \tilde \calD \uvec^c_{n+1}.
}
Here, it is clear that the matrix $I + \Delta t^2 \calH \tilde \calG \tilde \calD - 2 \rotrate \Delta t \: J_z \calR$ is sparse with banded-block-banded structure, and hence this system will be efficient to solve. Doing so at each iteration provides us with the coefficients vectors $\uvec^c_{n+1}, \tilde{\vec{h}}^c_{n+1}$ if we know $\uvec^c_{n}, \tilde{\vec{h}}^c_{n}$. Of course, when evaluating the function $h$, we can simply gather back the coefficients vector $\vec{h}^c$ by $\vec{h}^c = E_s \tilde{\vec{h}}^c$.

\bstodo{Images}
